The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-10-05 article.tex
Functional use of whitespace increasing readability

  Atom type perception is core cheminformations functionality: the
  atom types describe chemical features of atoms, such as the number
  of neighbors, possible formal charges, (approximate) hybridization,
  electron distribution over orbitals and so on. However, previous
  versions of the CDK implemented atom type perception as part of
  different algorithms, resulting in duplicated and sometimes
  divergent typing schemes. As a result it was cumbersome to add new
  atom types and implement support for new charged and radical species
  in a consistent manner.

@@

\begin{verbatim}
for (IAtom atom : molecule.atoms()) {
  type = matcher.findMatchingAtomType(molecule, atom);
}
\end{verbatim}

@@

\begin{verbatim}
types = matcher.findMatchingAtomTypes(molecule);
\end{verbatim}

@@

  \begin{itemize}
   \item perception from 2D coordinates;
   \item perception from 3D coordinates;
   \item wedge assignment;
   \item graph (sub)isomorphism matching;
   \item SMARTS matching; and
   \item canonicalization.
  \end{itemize}

@@

\begin{verbatim}
for (IAtom atom : molecule.atoms()) {
  String signature = new AtomSignature(atom, 2, molecule).toCanonicalString();
}
\end{verbatim}

@@

\begin{verbatim}
String signature = new MoleculeSignature(molecule).toCanonicalString();
\end{verbatim}

@@

\begin{verbatim}
new DepictionGenerator()
  .withSize(300, 300)
  .depict(benzene)
  .writeTo("BenzeneLocalized.png");
\end{verbatim}

@@

\begin{verbatim}
StructureDiagramGenerator sdg = new StructureDiagramGenerator();
sdg.setMolecule(someMolecule);
sdg.generateCoordinates();
IAtomContainer layedOutMol = sdg.getMolecule();
\end{verbatim}

@@

\begin{verbatim}
someMolecule = parser.parseSmiles("OC/C(C)=C/[C@](F)(Cl)Br");
sdg.setMolecule(someMolecule);
sdg.generateCoordinates();
IAtomContainer layedOutMol = sdg.getMolecule();
\end{verbatim}

@@

\begin{verbatim}
Isotopes ifac = Isotopes.getInstance();
MolecularFormulaRange range =
  new MolecularFormulaRange();
range.addIsotope( ifac.getMajorIsotope("C"), 8, 20);
range.addIsotope( ifac.getMajorIsotope("H"), 0, 20);
range.addIsotope( ifac.getMajorIsotope("O"), 0, 1);
range.addIsotope( ifac.getMajorIsotope("N"), 0, 1);

MolecularFormulaGenerator tool =
  new MolecularFormulaGenerator(
    SilentChemObjectBuilder.getInstance(),
    133.0, 133.1, range
  );
IMolecularFormulaSet mfSet = tool.getAllFormulas();
for (mf in mfSet) {
  println MolecularFormulaManipulator.getString(mf) + " " +
    MolecularFormulaManipulator.getTotalExactMass(mf)
}
\end{verbatim}

@@

\begin{verbatim}
C11H   133.007825032
C9H11N 133.089149352
C9H9O  133.065339908
C8H7NO 133.052763844
\end{verbatim}

@@

\begin{verbatim}
  SmilesParser parser = new SmilesParser(
    SilentChemObjectBuilder.getInstance()
  );
  parser.kekulise(false);
  mol = parser.parseSmiles(smi);
\end{verbatim}

@@

\begin{verbatim}
  uniqueGenerator = SmilesGenerator.unique()
  aromaticGenerator = SmilesGenerator.generic().aromatic()
  smiles = uniqueGenerator.createSMILES(mol)
  smiles2 = atomaticGenerator.createSMILES(mol)
\end{verbatim}

@@

  \begin{verbatim}
    // initialize SMARTS pattern API
    Pattern pattern = 
      SmartsPattern.findSubstructure("O=[C,N]aa[N,O;!H0]");

@@

    // lazily iterate over all unique atom matches as query atom
    // index bijection to atoms in 'mol', 'rxn' can also be used
    for (int[] m : pattern.matchAll(mol)
                          .uniqueAtoms()) {
       ...
    }
  \end{verbatim}

@@

\begin{verbatim}
  allCycles = Cycles.all(mol)
  relevantCycles = Cycles.relevant(mol)
  essentialCycles = Cycles.essential(mol)
  sssrCycles = Cycles.sssr(mol)
\end{verbatim}

@@

\begin{verbatim}
Aromaticity aromaticity = new Aromaticity(
  ElectronDonation.cdk(), Cycles.cdkAromaticSet()
);
\end{verbatim}

@@

\begin{verbatim}
Aromaticity aromaticity = new Aromaticity(
  ElectronDonation.cdk(), Cycles.all()
);
\end{verbatim}

@@

\begin{verbatim}
IChemObjectBuilder builder; // previously defined
IAtom atom = builder.newInstance(IAtom.class, "C");
\end{verbatim}

>>>>

  Atom type perception is core cheminformatics functionality: the
  atom types describe chemical features of atoms, such as the number
  of neighbors, possible formal charges, (approximate) hybridization,
  electron distribution over orbitals and so on. However, previous
  versions of the CDK implemented atom type perception as part of
  different algorithms, resulting in duplicated and sometimes
  divergent typing schemes. As a result it was cumbersome to add new
  atom types and implement support for new charged and radical species
  in a consistent manner.

@@

\vspace{0.2cm}
\begin{verbatim}
for (IAtom atom : molecule.atoms()) {
  type = matcher.findMatchingAtomType(molecule, atom);
}
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
types = matcher.findMatchingAtomTypes(molecule);
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
  \begin{itemize}
   \item perception from 2D coordinates;
   \item perception from 3D coordinates;
   \item wedge assignment;
   \item graph (sub)isomorphism matching;
   \item SMARTS matching; and
   \item canonicalization.
  \end{itemize}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
for (IAtom atom : molecule.atoms()) {
  String signature = new AtomSignature(atom, 2, molecule).toCanonicalString();
}
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
String signature = new MoleculeSignature(molecule).toCanonicalString();
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
new DepictionGenerator()
  .withSize(300, 300)
  .depict(benzene)
  .writeTo("BenzeneLocalized.png");
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
StructureDiagramGenerator sdg = new StructureDiagramGenerator();
sdg.setMolecule(someMolecule);
sdg.generateCoordinates();
IAtomContainer layedOutMol = sdg.getMolecule();
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
someMolecule = parser.parseSmiles("OC/C(C)=C/[C@](F)(Cl)Br");
sdg.setMolecule(someMolecule);
sdg.generateCoordinates();
IAtomContainer layedOutMol = sdg.getMolecule();
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
Isotopes ifac = Isotopes.getInstance();
MolecularFormulaRange range =
  new MolecularFormulaRange();
range.addIsotope( ifac.getMajorIsotope("C"), 8, 20);
range.addIsotope( ifac.getMajorIsotope("H"), 0, 20);
range.addIsotope( ifac.getMajorIsotope("O"), 0, 1);
range.addIsotope( ifac.getMajorIsotope("N"), 0, 1);

MolecularFormulaGenerator tool =
  new MolecularFormulaGenerator(
    SilentChemObjectBuilder.getInstance(),
    133.0, 133.1, range
  );
IMolecularFormulaSet mfSet = tool.getAllFormulas();
for (mf in mfSet) {
  println MolecularFormulaManipulator.getString(mf) + " " +
    MolecularFormulaManipulator.getTotalExactMass(mf)
}
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
C11H   133.007825032
C9H11N 133.089149352
C9H9O  133.065339908
C8H7NO 133.052763844
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
  SmilesParser parser = new SmilesParser(
    SilentChemObjectBuilder.getInstance()
  );
  parser.kekulise(false);
  mol = parser.parseSmiles(smi);
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
  uniqueGenerator = SmilesGenerator.unique()
  aromaticGenerator = SmilesGenerator.generic().aromatic()
  smiles = uniqueGenerator.createSMILES(mol)
  smiles2 = atomaticGenerator.createSMILES(mol)
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
  \begin{verbatim}
    // initialize SMARTS pattern API
    Pattern pattern = 
      SmartsPattern.findSubstructure("O=[C,N]aa[N,O;!H0]");

@@

    // lazily iterate over all unique atom matches as query atom
    // index bijection to atoms in 'mol', 'rxn' can also be used
    for (int[] m : pattern.matchAll(mol)
                          .uniqueAtoms()) {
       ...
    }
  \end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
  allCycles = Cycles.all(mol)
  relevantCycles = Cycles.relevant(mol)
  essentialCycles = Cycles.essential(mol)
  sssrCycles = Cycles.sssr(mol)
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
Aromaticity aromaticity = new Aromaticity(
  ElectronDonation.cdk(), Cycles.cdkAromaticSet()
);
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
Aromaticity aromaticity = new Aromaticity(
  ElectronDonation.cdk(), Cycles.all()
);
\end{verbatim}
\vspace{0.2cm}

@@

\vspace{0.2cm}
\begin{verbatim}
IChemObjectBuilder builder; // previously defined
IAtom atom = builder.newInstance(IAtom.class, "C");
\end{verbatim}
\vspace{0.2cm}
